title: Assignment_ML
author: "by Prakhar Gupta"
output:
  html_document:
    keep_md: true
    fig_height: 9
    fig_width: 9
---

### Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: * * * http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).


### Data

The training data for this project are available here:

* https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

* https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.


## Environment Preparation

Load all the required libraries

```{r, cache = T}
library(knitr)
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
library(corrplot)
library(RColorBrewer)
set.seed(1234)
```

### Read the Data
After downloading the data from the data source, read the two csv files 

```{r, cache = T}
training  <- 'pml-training.csv'
quizdata   <- 'pml-testing.csv'
train_data <- read.csv(training, strip.white = TRUE, na.strings = c("NA",""))
quiz_data  <- read.csv(quizdata,  strip.white = TRUE, na.strings = c("NA",""))
dim(train_data)
dim(quiz_data)
```

The training data set contains 19622 observations and 160 variables and testing data set contains 20 observations and 160 variables.

### partitioned the train data 

I’ll split the training data into training and testing partitions and use the pml-testing.csv as a validation sample

```{r, cache = T}
set.seed(127)
in_train  <- createDataPartition(train_data$classe, p=0.75, list=FALSE)
train_set <- train_data[ in_train, ]
test_set  <- train_data[-in_train, ]
dim(train_set)
dim(test_set)
```

### Clean Data 

There are  plenty of NA, that can be removed with the cleaning procedures below. The Near Zero variance (NZV) variables are also removed and the ID variables as well.

```{r, cache=T}
nzv_var <- nearZeroVar(train_set)
train_set <- train_set[ , -nzv_var]
test_set  <- test_set [ , -nzv_var]
dim(train_set)
dim(test_set)
```
## Remove variables that are NA. A threshlod of 95 % is selected.

```{r, cache=T}
allna    <- sapply(train_set, function(x) mean(is.na(x))) > 0.95
train_set <- train_set[, allna==FALSE]
test_set  <- test_set[, allna==FALSE]
dim(train_set)
dim(test_set)
```

##remove identification  variables (columns 1 to 5)

```{r, cache=T}
train_set <- train_set[, -(1:5)]
test_set  <- test_set[, -(1:5)]
dim(train_set)
dim(test_set)
```

As per the cleaning process above, the number of variables for the analysis has been reduced to 54 only.


### Correlation Analysis

A correlation matrix among variables.Select “FPC” for the first principal component order. 

```{r, cache=T}
corrmatrix <-cor(train_set[, -length(names(train_set))])
corrplot(corrmatrix, order = "FPC", method = "color", type = "lower", 
         tl.cex = 0.8, tl.col = rgb(0, 0, 0))
```


The highly correlated variables are shown in dark colors in the graph above.The  highly correlated their colors are either dark blue (for a positive correlation) or dark red (for a negative corraltions).

### Prediction Model 

Three methods will be applied to model the regressions (in the Train dataset) and the best one (with higher accuracy when applied to the Quiz dataset) will be used for the quiz predictions. 

The methods are: 
* Random Forests, 
* Decision Tree and 
* Generalized Boosted Model described below.

A Confusion Matrix is plotted at the end of each analysis for better visualize the accuracy of the models.

## A)Random Forest

* use Random Forest method with Cross Validation, 3 folds
* number =3 is # folds of the training data
* Cross Validation achieved with trainControl method set to “cv”
* ntree =250 is the number of tree 

```{r, cache=T}
#model fit
set.seed(1234)
controlRF <- trainControl(method="cv", number=3, verboseIter=FALSE)
modelRf  <- train(classe ~ ., data=train_set, method="rf",
                          trControl=controlRF,ntree=250)
modelRf$finalModel
# prediction on Test dataset
#Check the accuracy of the model by comparing the predictions to the actual results
predictRf <- predict(modelRf, newdata=test_set)
confMatRandForest <- confusionMatrix(predictRf, test_set$classe)
confMatRandForest
accuracy <- postResample(predictRf, test_set$classe)
accuracy
outof_sample_error <- 1 - as.numeric(confusionMatrix(test_set$classe, predictRf)$overall[1])
outof_sample_error
#plot matrix results
plot(confMatRandForest$table, col = confMatRandForest$byClass, 
     main = paste("Random Forest - Accuracy =",
                  round(confMatRandForest$overall['Accuracy'], 4)))
```
